---
title: "2. Exploratory Data Analysis. Ejercicios"
date: "Febrero 2025"
output: 
  html_document:
    toc: true
    toc_depth: 1
    theme: cerulean
---


<style>
body {
text-align: justify}
</style>

<style type="text/css">
body{/*Normal*/
     font-size: 20px;
}
</style>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(discrtr) # A companion package for the book Introduction to Discrete Choice Analysis with `R`
library(mlogit) # Multinomial Logit Models
library(readr) # Read Rectangular Text Data
library(dplyr)
library(gplots) 
library(ggmosaic) 
library(treemapify)
library(ggridges)
library(ggalluvial)
library(plotly)
library(ggalluvial)
library(ggplot2)
```


Call data set Car from package mlogit. Once you have loaded the data set, do the following:


```{r}
data("Car",package = "mlogit")
head(Car)
```


* How many variables are there in this data set and of which type (i.e., categorical/quantitative)?

```{r}
glimpse(Car)
```


* Choose four relevant categorical variables from this data set and describe them using univariate, bivariate, and multivariate techniques. Discuss your results.

```{r}
ggplot(Car)+
  geom_bar(aes(type3, fill=type3), color="black", show.legend = FALSE)+
  theme_bw()
```

```{r}
ggplot(Car)+
  geom_bar(aes(fuel5, fill=fuel5 ), color="black", show.legend = FALSE)+
  theme_bw()
```

```{r}
ggplot(Car)+
  geom_bar(aes(type4, fill=type4 ), color="black", show.legend = FALSE)+
  theme_bw()
```

```{r}
ggplot(Car)+
  geom_bar(aes(type5, fill=type5 ), color="black", show.legend = FALSE)+
  theme_bw()
```

* What ideas about individuals’ choices regarding car do you develop from this EDA exercise?

```{r}
ggplot(Car)+
  geom_density(aes(type3, fill="type3"), color="black", alpha=0.5)+
  geom_density(aes(fuel5 , fill="fuel5"), color="black", alpha=0.5)+
  geom_density(aes(type4, fill="type4"), color="black", alpha=0.5)+
  geom_density(aes(type5, fill="type5"), color="black", alpha=0.5)+
  scale_fill_manual(values=c("type3"= "firebrick", "fuel5"="yellow", "type4"="darkgoldenrod2", "type5"="cyan"), name="Mode")+
  xlab("Cost")+
theme_bw()
  
```
Se carga el conjunto de datos, se muestran las variables y su tipo:
```{r}
Car_alluvia <- Car |> 
  select(choice, type1, type2, fuel1, fuel2) |> 
  group_by(choice, type1, type2, fuel1, fuel2) |> 
  summarize(frequency = n(), .groups = "drop")
```


```{r}
Car_alluvia |> 
  ggplot(aes(y = frequency, axis1 = choice, axis2 = type1, axis3 = type2, axis4 = fuel1, axis5 = fuel2)) +
  geom_alluvium(aes(fill = choice), width = 1/3, color = "black") +
  geom_stratum(width = 1/3, fill = "black", color = "grey") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), color = "white", size = 3) +
  scale_x_discrete(limits = c("Choice", "Tipo 1", "Tipo 2", "Combustible 1", "Combustible 2"), expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Spectral") + 
  theme_minimal()
```



```{r}
ggplot(data = Car_alluvia, 
       aes(axis1 = type2, axis2 = fuel1, axis3 = type1, axis4 = fuel2, y = frequency)) +
  geom_alluvium(aes(fill = type2), width = 1/3) +
  geom_stratum(width = 1/3, fill = "black", color = "grey") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), color = "white", size = 3) +
  scale_x_discrete(labels = c("Tipo 3", "Combustible 5", "Tipo 4", "Tipo 5"),
                   expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Spectral") + 
  theme_minimal()

```
```{r}
data(Car)
glimpse(Car)
```




En este caso es necesario añadir columnas acerca de la elección:


```{r}
Car_alluvia <- Car |> 
  select(choice, type1, type2, fuel1, fuel2) |> 
  group_by(choice, type1, type2, fuel1, fuel2) |> 
  summarize(frequency = n(), .groups = "drop")

```
```{r}
summary(Car_alluvia)
```

