---
title: "Capitulo 7 Ejercicios 2,3 y 4"
author: "Gabriel Omar Maya Briseño"
date: "2025-05-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(discrtr) # A companion package for the book Introduction to Discrete Choice Analysis with `R`
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(mlogit) # Multinomial Logit Models
library(readr) # Read Rectangular Text Data
library(gplots) # Various R Programming Tools for Plotting Data
library(ggmosaic) # Mosaic Plots in the 'ggplot2' Framework
library(treemapify)
library(ggridges)
library(ggalluvial)
library(evd)
library(htmlwidgets) # HTML Widgets for R
library(kableExtra) # Construct Complex Table with kable and Pipe Syntax
library(plotly) # Create interactive web graphics
library(tidyr) # Tidy messy data

library(evd)
```

# introduccion

```{r}
data("Heating", package = "mlogit")
H <- mlogit.data(Heating, shape = "wide", choice = "depvar", varying = c(3:12))
model3 <- mlogit(depvar ~ ic + oc, Heating, shape = "wide", choice = "depvar", reflevel = "ec", varying = c(3:12))
summary(model3)
```

```{r}
X <- model.matrix(model3)
head(X)
```

```{r}
alt <- index(H)$alt
Xmec <- X[alt != "ec",]
Xmer <- X[alt != "er",]
Xmgc <- X[alt != "gc",]
Xmgr <- X[alt != "gr",]
Xmhp <- X[alt != "hp",]
```

```{r}
# Unique identifiers by decision-maker
chid <- index(H)$chid
# Remove the fifth identifier for each decision-maker
chid <- chid[-seq(1, length(chid), 5)]
```

```{r}
# After removing ec
exp_Xb_mec <- as.numeric(exp(Xmec %*% coef(model3)))
sum_exp_Xb_mec <- as.numeric(tapply(exp_Xb_mec, sort(chid), sum))
P_mec <- exp_Xb_mec / sum_exp_Xb_mec[sort(chid)]
# After removing er
exp_Xb_mer <- as.numeric(exp(Xmer %*% coef(model3)))
sum_exp_Xb_mer <- as.numeric(tapply(exp_Xb_mer, sort(chid), sum))
P_mer <- exp_Xb_mer / sum_exp_Xb_mer[sort(chid)]
# After removing gc
exp_Xb_mgc <- as.numeric(exp(Xmgc %*% coef(model3)))
sum_exp_Xb_mgc <- as.numeric(tapply(exp_Xb_mgc, sort(chid), sum))
P_mgc <- exp_Xb_mgc / sum_exp_Xb_mgc[sort(chid)]
# After removing gr
exp_Xb_mgr <- as.numeric(exp(Xmgr %*% coef(model3)))
sum_exp_Xb_mgr <- as.numeric(tapply(exp_Xb_mgr, sort(chid), sum))
P_mgr <- exp_Xb_mgr / sum_exp_Xb_mgr[sort(chid)]
# After removing hp
exp_Xb_mhp <- as.numeric(exp(Xmhp %*% coef(model3)))
sum_exp_Xb_mhp <- as.numeric(tapply(exp_Xb_mhp, sort(chid), sum))
P_mhp <- exp_Xb_mhp / sum_exp_Xb_mhp[sort(chid)]
```

```{r}
# After removing ec
P_mec <- data.frame(matrix(P_mec, ncol = 4, byrow = TRUE))
P_mec <- transmute(P_mec,
# Remove this alternative from the choice set
ec = NA, er = P_mec[, 1], gc = P_mec[, 2], gr = P_mec[, 3], hp = P_mec[, 4])

# After removing er
P_mer <- data.frame(matrix(P_mer, ncol = 4, byrow = TRUE))
P_mer <- transmute(P_mer, ec = P_mer[, 1],
# Remove this alternative from the choice set
er = NA, gc = P_mer[, 2], gr = P_mer[, 3], hp = P_mer[, 4])

# After removing gc
P_mgc <- data.frame(matrix(P_mgc, ncol = 4, byrow = TRUE))
P_mgc <- transmute(P_mgc, ec = P_mgc[, 1], er = P_mgc[, 2],
# Remove this alternative from the choice set
gc = NA, gr = P_mgc[, 3], hp = P_mgc[, 4])

# After removing gr
P_mgr <- data.frame(matrix(P_mgr, ncol = 4, byrow = TRUE))
P_mgr <- transmute(P_mgr, ec = P_mgr[, 1], er = P_mgr[, 2], gc = P_mgr[, 3],
# Remove this alternative from the choice set
gr = NA, hp = P_mgr[, 4])

# After removing hp
P_mhp <- data.frame(matrix(P_mhp, ncol = 4, byrow = TRUE))
P_mhp <- transmute(P_mhp, ec = P_mhp[, 1], er = P_mhp[, 2], gc = P_mhp[, 3], gr = P_mhp[, 4],
# Remove this alternative from the choice set
hp = NA)
```

# tabla


```{r}
df <- data.frame(Alternative = c("None", "ec", "er", "gc", "gr", "hp" ),
rbind(apply(fitted(model3,
outcome = FALSE),
2, mean),
apply(P_mec, 2, mean),
apply(P_mer, 2, mean),
apply(P_mgc, 2, mean),
apply(P_mgr, 2, mean),
apply(P_mhp, 2, mean))
)
df %>%
kable(col.names = c("Alternative Removed",
"ec",
"er",
"gc",
"gr",
"hp"),
digits = 3) %>%
kable_styling()
```




# 2. Use model nl2 in this chapter and calculate the   direct-point elasticity at the mean values of the variables, for an increase in the installation costs of Gas Central systems.

Primero se copian los primeros 5 elementos de la matriz del modelo (una fila por cada alternativa)

```{r}
nl2 <- mlogit(depvar ~ ic + oc, H, nests = list(room = c( 'er', 'gr'), central = c('ec', 'gc', 'hp')),
              un.nest.el = TRUE, steptol = 1e-12)

summary(nl2)
```


Elasticidad

```{r}
X_mean <- model.matrix(nl2)[1:5,]
alt <- index(H)$alt[1:5]
```

A continuación se calculan las medias de los costos de instalación y operación para cada sistema

```{r}
mean_ic <- H %>% group_by(alt) %>% summarize(ic = mean(ic)) %>% arrange(alt)

mean_oc <- H %>% group_by(alt) %>% summarize(oc = mean(oc)) %>% arrange(alt)
```

Se calcula la exponencial de las funciones de utilidad con los costos dados por las medias para cada alternativa; nótese que se utilizó el coeficiente del valor inclusivo

```{r}
# Electric central
exp_V_ec <- exp((X_mean[alt == c("ec"), "oc"] * coef(nl2)["oc"] +
            X_mean[alt == c("ec"), "ic"] * coef(nl2)["ic"]) / coef(nl2)["iv"])

# Electric room
exp_V_er <- exp((coef(nl2)["(Intercept):er"] + X_mean[alt == c("er"), "oc"] * coef(nl2)["oc"] +
            X_mean[alt == c("er"), "ic"] * coef(nl2)["ic"]) / coef(nl2)["iv"])

# Gas central
exp_V_gc <- exp((coef(nl2)["(Intercept):gc"] + X_mean[alt == c("gc"), "oc"] * coef(nl2)["oc"] +
            X_mean[alt == c("gc"), "ic"] * coef(nl2)["ic"]) / coef(nl2)["iv"])

# Gas room
exp_V_gr <- exp((coef(nl2)["(Intercept):gr"] + X_mean[alt == c("gr"), "oc"] * coef(nl2)["oc"] +
            X_mean[alt == c("gr"), "ic"] * coef(nl2)["ic"]) / coef(nl2)["iv"])

# Heat pump
exp_V_hp <- exp((coef(nl2)["(Intercept):hp"] + X_mean[alt == c("hp"), "oc"] * coef(nl2)["oc"] +
            X_mean[alt == c("hp"), "ic"] * coef(nl2)["ic"]) / coef(nl2)["iv"])
```

```{r}
# Conditional probabilities of systems within the central nest
cp_c <- data.frame(ec = exp_V_ec / (exp_V_ec + exp_V_gc + exp_V_hp),
                   gc = exp_V_gc / (exp_V_ec + exp_V_gc + exp_V_hp),
                   hp = exp_V_hp / (exp_V_ec + exp_V_gc + exp_V_hp))

# Conditional probabilities of systems within the room nest
cp_r <- data.frame(er = exp_V_er / (exp_V_er + exp_V_gr),
                   gr = exp_V_gr / (exp_V_er + exp_V_gr))
```

```{r}
#After removing ec
mp <- data.frame(central = exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_gc + exp_V_hp))
                          / (exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_gc + exp_V_hp)) +
                          exp((coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)))),
                room = exp(coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)) / (exp(coef(nl2)["iv"] * log(exp_V_gc + exp_V_hp)) +
                      exp((coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)))))
```

```{r}
nlp <- data.frame(system = c("ec", "er", "gc", "gr", "hp"),
                  # Conditional probability
                  cp = c(cp_c$ec, cp_r$er, cp_c$gc, cp_r$gr, cp_c$hp),
                  # Marginal probability
                  mp = c(mp$central, mp$room, mp$central, mp$room, mp$central),
                  beta_ic = c(as.numeric(nl2$coefficients["ic"])),
                  beta_oc = c(as.numeric(nl2$coefficients["oc"])),
                  lambda = c(as.numeric(nl2$coefficients["iv"]))) %>%
# Joint probability
mutate(p = cp * mp)
```

# 3. Use model nl2 in this chapter and calculate the cross-point elasticity at the mean values of the variables, for a 1% increase in the operation costs of Gas Central systems.

```{r}
nlp <- cbind(nlp, X_mean[,5:6]) %>%
# Increase installation cost 1%
mutate(ic_1pct = 1.01 * ic)
```

```{r}
direct_elasticities <- nlp %>%
                       transmute(DEM = ((1 - mp) + (1 - cp) * (1 - lambda)/lambda) * beta_ic * ic)

direct_elasticities
```

# 4. Re-estimate the nested logit model in this chapter, but change the nests to types of energy as follows: • Gas: gas central, gas room. • Electricity: electric central, electric room, heat pump. Useasinglecoefficientfortheinclusivevariables(i.e.,set un.nest.el = TRUE). Are the results reasonable? Discuss.

```{r}
nl1 <- mlogit(depvar ~ oc + ic, H, nests = list(room=c('er', 'gr'), central = c('ec', 'gc', 'hp')), steptol=1e-12)

summary(nl1)
```

```{r}
1-nl1$coefficients["iv:room"]
```
```{r}
1-nl1$coefficients["iv:central"]
```
Efectivamente, las correlaciones en los nidos son muy altas. Podemos determinar cuándo la correlación es significativa por medio del t-test

```{r}
(nl1$coefficients["iv:room"] - 1) / sqrt(vcov(nl1)["iv:room","iv:room"])
```

```{r}
(nl1$coefficients["iv:central"] - 1) / sqrt(vcov(nl1)["iv:central","iv:central"])
```

```{r}
lrtest(model3, nl1)
```
```{r}
nl2 <- mlogit(depvar ~ ic + oc, H, nests = list(room = c( 'er', 'gr'), central = c('ec', 'gc', 'hp')),
              un.nest.el = TRUE, steptol = 1e-12)

summary(nl2)
```

```{r}
lrtest(nl2, nl1)
```
cambie los nidos a tipos de energía como se indica a continuación: • Gas: central de gas, habitación de gas. • Electricidad: central eléctrica, habitación eléctrica, bomba de calor.

```{r}
nl_3 <- mlogit(depvar ~ oc + ic, H, nests = list(electric=c('ec', 'er', 'hp'), gas = c('gc', 'gr')), steptol=1e-12)

#summary(nl_3)

nl_3$coefficients
```
Patrones de sustitución con el modelo logit anidado

```{r}
X <- model.matrix(nl2)
head(X,12)
```

```{r}
# Electric central
exp_V_ec <- exp((X[alt == c("ec"), "oc"] * coef(nl2)["oc"] + X[alt == c("ec"), "ic"] * coef(nl2)["ic"]) / coef(nl2)["iv"])

# Gas central
exp_V_gc <- exp((coef(nl2)["(Intercept):gc"] + X[alt == c("gc"), "oc"] * coef(nl2)["oc"] + X[alt == c("gc"), "ic"] * 
                coef(nl2)["ic"])/ coef(nl2)["iv"])

# Heat pump
exp_V_hp <- exp((coef(nl2)["(Intercept):hp"] + X[alt == c("hp"), "oc"] * coef(nl2)["oc"] + X[alt == c("hp"), "ic"] * 
                   coef(nl2)["ic"])/ coef(nl2)["iv"])

# Electric room
exp_V_er <- exp((coef(nl2)["(Intercept):er"] + X[alt == c("er"), "oc"] * coef(nl2)["oc"] + X[alt == c("er"), "ic"] *
                 coef(nl2)["ic"])/ coef(nl2)["iv"])

# Gas room
exp_V_gr <- exp((coef(nl2)["(Intercept):gr"] + X[alt == c("gr"), "oc"] * coef(nl2)["oc"] + X[alt == c("gr"), "ic"] *
                 coef(nl2)["ic"]) / coef(nl2)["iv"])
```

Las probabilidades condicionales son los modelos logit en cada nido

```{r}
# Conditional probabilities of systems within the central nest after removing ec
cp_mec_c <- data.frame(gc = exp_V_gc / (exp_V_gc + exp_V_hp), hp = exp_V_hp / (exp_V_gc + exp_V_hp))

# Conditional probabilities of systems within the central nest after removing gc
cp_mgc_c <- data.frame(ec = exp_V_ec / (exp_V_ec + exp_V_hp), hp = exp_V_hp / (exp_V_ec + exp_V_hp))

# Conditional probabilities of systems within the central nest  after removing hp
cp_mhp_c <- data.frame(ec = exp_V_ec / (exp_V_ec + exp_V_gc), gc = exp_V_gc / (exp_V_ec + exp_V_gc))

# Conditional probabilities of systems within the room nest after removing a system in the central nest
cp_mc_r <- data.frame(er = exp_V_er / (exp_V_er + exp_V_gr), gr = exp_V_gr / (exp_V_er + exp_V_gr))

# Conditional probabilities of systems within the room nest after removing er
cp_mer_r <- data.frame(gr = exp_V_gr / (exp_V_gr))

# Conditional probabilities of systems within the room nest after removing gr
cp_mgr_r <- data.frame(er = exp_V_er / (exp_V_er))

# Conditional probabilities of systems within the central nest after removing a system in the room nest
cp_mr_c <- data.frame(ec = exp_V_ec / (exp_V_ec + exp_V_gc + exp_V_hp), gc = exp_V_gc / (exp_V_ec + exp_V_gc + exp_V_hp),
                      hp = exp_V_hp / (exp_V_ec + exp_V_gc + exp_V_hp))
```

Las probabilidades marginales son las probabilidades logit de escoger un nido dada la utilidad máxima esperada de cada nido

```{r}
#After removing ec
mp_mec <- data.frame(central = exp(coef(nl2)["iv"] * log(exp_V_gc + exp_V_hp)) / (exp(coef(nl2)["iv"] 
                               * log(exp_V_gc + exp_V_hp)) + exp((coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)))),
                     room = exp(coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)) / (exp(coef(nl2)["iv"] * log(exp_V_gc + exp_V_hp))
                            + exp((coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)))) )

#After removing gc
mp_mgc <- data.frame(central = exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_hp)) / (exp(coef(nl2)["iv"] * log(exp_V_ec + 
                               exp_V_hp)) + exp((coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)))),
                     room = exp(coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)) / (exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_hp))                             + exp((coef(nl2)["iv"] * log(exp_V_er + exp_V_gr))))  )

#After removing hp
mp_mhp <- data.frame(central = exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_gc)) / (exp(coef(nl2)["iv"] * log(exp_V_ec +
                               exp_V_gc)) + exp((coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)))),
                     room = exp(coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)) /
                            (exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_gc)) + exp((coef(nl2)["iv"] * log(exp_V_er + exp_V_gr)))) )

#After removing er
mp_mer <- data.frame(central = exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_gc + exp_V_hp)) / (exp(coef(nl2)["iv"] *
                               log(exp_V_ec + exp_V_gc + exp_V_hp)) + exp((coef(nl2)["iv"] * log(exp_V_gr)))),
                     room = exp(coef(nl2)["iv"] * log(exp_V_gr)) / (exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_gc + exp_V_hp))
                            + exp((coef(nl2)["iv"] * log(exp_V_gr))))  )

#After removing gr
mp_mgr <- data.frame(central = exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_gc + exp_V_hp)) / (exp(coef(nl2)["iv"] *
                               log(exp_V_ec + exp_V_gc + exp_V_hp)) + exp((coef(nl2)["iv"] * log(exp_V_er)))),
          room = exp(coef(nl2)["iv"] * log(exp_V_er)) / (exp(coef(nl2)["iv"] * log(exp_V_ec + exp_V_gc + exp_V_hp)) + exp((coef(nl2)["iv"] * log(exp_V_er)))) )
```

Una vez que las probabilidades condicionales y marginales para cada caso han sido calculadas, las probabilidades de elección son el producto de dichas probabilidades

```{r}
#After removing ec
nlp_mec <- data.frame(cp_mec_c, cp_mc_r, mp_mec) %>% 
           transmute(p_ec = NA, p_gc = gc * central, p_hp = hp * central, p_er = er * room, p_gr = gr * room)

#After removing gc
nlp_mgc <- data.frame(cp_mgc_c, cp_mc_r, mp_mgc) %>%
           transmute(p_ec = ec * central, p_gc = NA, p_hp = hp * central, p_er = er * room, p_gr = gr * room)

#After removing hp
nlp_mhp <- data.frame(cp_mhp_c, cp_mc_r, mp_mhp) %>%
           transmute(p_ec = ec * central, p_gc = gc * central, p_hp = NA, p_er = er * room, p_gr = gr * room)

#After removing er
nlp_mer <- data.frame(cp_mr_c, cp_mer_r, mp_mer) %>%
           transmute(p_ec = ec * central, p_gc = gc * central, p_hp = hp * central, p_er = NA, p_gr = gr * room)

#After removing gr
nlp_mgr <- data.frame(cp_mr_c, cp_mgr_r, mp_mgr) %>%
           transmute(p_ec = ec * central, p_gc = gc * central, p_hp = hp * central, p_er = er * room, p_gr = NA)
```

Se realiza una revisión preliminar para verificar que la suma de las probabilidades para cada decisor es igual a uno

```{r}
summary(rowSums(nlp_mec, na.rm = TRUE))
```
```{r}
summary(rowSums(nlp_mgc, na.rm = TRUE))
```

```{r}
summary(rowSums(nlp_mhp, na.rm = TRUE))
```
```{r}
summary(rowSums(nlp_mer, na.rm = TRUE))
```
```{r}
summary(rowSums(nlp_mgr, na.rm = TRUE))
```
tabla 

```{r}
# Original adoption rates
# Using the fitted function to calculate the probabilities for each household
p_o <- apply(fitted(nl2, outcome = FALSE), 2, mean)

df <- data.frame(Alternative = c("None", "ec", "gc", "hp", "er", "gr" ),
                 rbind(c(p_o["ec"], p_o["gc"], p_o["hp"], p_o["er"], p_o["gr"]),
                       apply(nlp_mec, 2, mean), apply(nlp_mgc, 2, mean),
                       apply(nlp_mhp, 2, mean), apply(nlp_mer, 2, mean),
                       apply(nlp_mgr, 2, mean)))

df %>% kable(col.names = c("Alternative Removed", "ec", "gc", "hp", "er", "gr"),
             digits = 3) %>% kable_styling()
```

# ¿Son razonables los resultados?

Sí, los resultados son razonables, y el modelo está bien especificado tanto en su estructura como en la implementación. Tus pasos en R son sólidos y reflejan un buen entendimiento del modelo logit anidado. Si deseas profundizar aún más, podrías estimar efectos de política (como eliminar subsidios) o introducir variables adicionales en las utilidades.